Subject-specific curves for time series forecasting of smart meter demand

### 2019-11-14 sscts-paper-1.Rmd
Update SSCTS. Proofread.

@@

Several papers have examined forecasting electricity demand for buildings by fitting separate models to each building [@Ghofrani2011-tb; @Gajowniczek2014-ek; @Arora2016-zh; @Ben_Taieb2016-wl]. While some have attempted to improve forecasts by leveraging the hierarchical nature of electricity demand [@Ben_Taieb2017-it; @Ben_Taieb2016-ok] few, if any, have explored improving forecast accuracy using a mixed effects framework. If buildings behave in a similar manner a well-specified mixed model may produce more accurate forecasts than individual models. Furthermore, a mixed-model framework allows us to quantify differences between buildings which would not otherwise be possible when using a "building-specific" modelling approach. A mixed effects approach opens the door to scenario analyses by allowing us to estimate how demand might change under different equipment or usage scenarios.

@@

We have time series and attribute data for `r n_buildings` commercial office buildings located across Australia. We focus on business days only in our analysis as these are significantly more important than non-business days for energy management. Non-business days typically have far less demand than business days as equipment is non-operational. Note that our approach can be applied to non-business days as well.

@@

The simplest benchmark is a naive forecasting model, where the previous day's values are used. This is often a surprisingly effective forecasting approach [@Hyndman2018-ok]. The demand for a building at time $t$ is given by
$$
  y_{t} = y_{t-24 \text{ hours}} + \epsilon_{t}, \quad \epsilon_{t} \sim N(0,\sigma^2).
$$
Note that since we have restricted ourselves to business days only, $t-24 \text{ hours}$ is a slight abuse of notation and is used to represent the observed values from the last _business_ day. So a forecast for Monday will use observed values from the previous Friday. Using observed values from Sunday would produce a much weaker benchmark due to different demand dynamics on working and non-working days.

@@

A simple benchmark model is created by fitting a linear regression model to each building and period of the day. The demand for a building at time $t$ is given by
$$
\log y_{t} = \beta_{0,p} + \beta_{1,p} w_{0,t} + \epsilon_{t}, \quad \epsilon_{t} \sim N(0,\sigma^2_{p}),
$$
where $p$ is the 15-minute period of the day at time $t$, $w_{0,t}$ is the scaled temperature experienced^[The 0 subscript denotes no lag and is included to keep notation consistent with Table \@ref(tab:weather-vars).] at time $t$ and $\epsilon_{t}$ is the residual. We call these our "Individual Linear Regression" (ILR) models.

@@

\begin{table}[t]
\caption{\label{tab:weather-vars} Predictor variables evaluated during feature selection stage when determining $\mathcal{P}_t$.}

@@

We do not estimate a separate smooth relationship between weather variables and demand for each building in the pooled model. Instead, we estimate the population's relationship. So, for the $i$^th^ predictor we construct a smooth function $f_{i,p}$ for all buildings instead of a set of smooth functions $f_{b,i,l}$ for each building.

@@

In the notation that follows we include subscripts $b$ to denote each building. Unlike the individual models above, which had separate models fit to each building, all buildings are used when training the mixed effects models and so we include an additional subscript to denote this. Again, to capture changing demand characteristics across the day, separate models are fit for each 15-minute period of the day giving 96 models for each mixed effects formulation.

@@

We use the pooled linear regression model for feature selection as it is quick to fit using ordinary least squares and allows us to model buildings by using a dummy variable for each. Using a linear model also has the advantage of allowing us to efficiently compute the leave-one-out cross-validation (LOOCV) scores [@Seber2012-gu]. Training data is comprised of business days within a window of `r fcst_train_window` days prior to the month we wish to select variables for. As our experimental setup for the validation phase involves one-day ahead forecasts, this variable selection prevents us from using any data from the future. For example, when forecasting for any date in January, only data from the months of October, November and December would have been used to select predictors.

We use best subset selection during feature selection. So, given $p$ predictors we choose the combination of these that produce the best $R^2$ scores. Once the best model based on $R^2$ has been determined for each set of $p$ predictors, we use the LOOCV score to determine the best. The LOOCV score is chosen as it gives an estimate of the out of sample performance of our models. Figure \@ref(fig:fs-model-selection) shows the LOOCV scores for each predictor set of size $p$. During feature selection we chose to avoid greedy approaches such as forward or backward stepwise selection; or approaches that work systematically through lagged weather variables [@Hyndman2010-ui]. Naturally, greedy methods have computational benefits, but it is interesting to observe which features are chosen when _all_ possible predictor combinations are assessed.

@@

It is difficult to observe clear differences between the extended summer and winter periods due to the amount of variance in the chosen variables. However, the maximum temperature over the last 24 hours seems to be more consistently selected during business hours in summer. This seems reasonable as high temperatures in summer will result in additional office cooling.

@@

(ref:fs-alpha) Feature selection for summer and winter including shoulder months.

@@

Table \@ref(tab:accuracy-results) shows the MAE, MAPE, sMAPE and MASE for each model across the entire day, during business hours (7:00 am to 7:00 pm) and during non-business hours (7:00 pm to 7:00 am). The INS models have the worst forecasting accuracy across all metrics. This is likely due to overfitting - the natural splines are too flexible for some of the buildings with the amount of data used for training. Evidence of this can be seen in Figure \@ref(fig:one-day-ahead-forecast) where erratic spikes in the predicted demand can be seen. When moving to a mixed effects model that uses natural splines the predictive accuracy is greatly improved. The SSC and SSCATTR models consistently outperform all other models.

@@

(ref:one-day-ahead-forecast-cap) One-day ahead forecasts for `r format(p_one_day_ahead_date, "%d %B, %Y")`.

```{r one-day-ahead-forecast, fig.cap="(ref:one-day-ahead-forecast-cap)", fig.height=7*1.4}
model_df %>%
  filter(fcst_date == p_one_day_ahead_date) %>%
  unnest(preds) %>%
  # filter(bid %in% sample(unique(bid), 3),
  #        fcst_date %in% (dmy("13/3/2017")+0:4)) %>%
  filter(bid %in% sample(unique(bid), 6)) %>%
  mutate(ts = as_datetime(fcst_date) + minutes((period - 1) * 24 * 60 / 96)) %>%
  ggplot(aes(x = ts)) +
  geom_line(aes(y = wh), colour = "black", linetype = "dashed") +
  geom_line(aes(y = pred, colour = model), alpha = 0.7) +
    facet_wrap(~bid, ncol = 2) +
  labs(
    y = expression(Normalised ~ electricity ~ (Wh / m^2)),
    x = "Date",
    colour = "Model"
  )
```

@@

Having confirmed that our mixed models produce satisfactory predictions compared to individual provides us with justification for using them for scenario analysis. We can investigate how changing certain variable values increases or decreases energy demand in buildings. We fit the SSCATTR model on all training data before producing predictions. This is an ex-post forecasting scenario since we are using known data for our predictor variables when predicting demand.

@@

We fit models to `r n_buildings` buildings across Australia. Separate models for each building were fitted as a benchmark. The overall predictive power of several random effects model and mixed effects model were assessed against this benchmark. One-day ahead forecasts were produced for business days over a year using all forecast methods. Based on the MAE, MAPE, sMAPE and MASE scores of each model the SSCATTR model performed best. We concluded that moving towards a mixed model approach improved predictive power and also had the additional advantage of allowing us to include fixed effects which could be used for scenario analyses.

@@

```{r fs-all-periods-months, fig.cap="(ref:fs-all-periods-months)", fig.height = 7*1.4}
fs_df %>%
  select(period, month, vars) %>%
  unnest() %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(
    selected = tidyr::replace_na(selected, FALSE),
    vars = fct_rev(factor(vars, levels = predictor_vars)),
    predictor_type = case_when(
      str_detect(vars, "^scaled_temperature$") ~ "Forced variable",
      str_detect(vars, "^scaled_temperature_lag") ~ "Temperature lags",
      str_detect(vars, "^scaled_temperature_(?!lag)") ~ "Temperature statistic",
      str_detect(vars, "^scaled_wh_lag") ~ "Demand lags",
      TRUE ~ "Other"
    )
  ) %>%
  filter(selected == TRUE) %>%
  ggplot(aes(x = period, y = vars, fill = predictor_type)) +
  geom_tile() +
  facet_wrap(~month) +
  # scale_fill_manual(values = c("white", "#3d6b78")) +
  labs(
    x = "Period",
    y = "Predictor",
    fill = "Predictor type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(8, 96, 8))
```
\newpage

>>>>

set.seed(123412)

@@

Several papers have examined forecasting electricity demand for buildings by fitting separate models to each building [@Ghofrani2011-tb; @Gajowniczek2014-ek; @Arora2016-zh; @Ben_Taieb2016-wl]. While some have attempted to improve forecasts by leveraging the hierarchical nature of electricity demand [@Ben_Taieb2017-it; @Ben_Taieb2017-ok] few, if any, have explored improving forecast accuracy using a mixed effects framework. If buildings behave in a similar manner a well-specified mixed model may produce more accurate forecasts than individual models. Furthermore, a mixed-model framework allows us to quantify differences between buildings which would not otherwise be possible when using a "building-specific" modelling approach. A mixed effects approach opens the door to scenario analyses by allowing us to estimate how demand might change under different equipment or usage scenarios.

@@

We have time series and attribute data for `r n_buildings` commercial office buildings located across Australia. We focus on business days in our analysis as these are significantly more important than non-business days for energy management. Non-business days typically have far less demand than business days as equipment is non-operational. Note that our approach can be applied to non-business days as well.

@@

The simplest benchmark is a naive forecasting model, where the previous day's values are used. This is often a surprisingly effective forecasting approach [@Hyndman2018-ok]. The demand for a building at time $t$ is given by
$$
  y_{t} = y_{t-24 \text{ hours}} + \epsilon_{t}, \quad \epsilon_{t} \sim N(0,\sigma^2).
$$
Note that since we have restricted ourselves to business days, $t-24 \text{ hours}$ is a slight abuse of notation and is used to represent the observed values from the last _business_ day. So a forecast for Monday will use observed values from the previous Friday. Using observed values from Sunday would produce a much weaker benchmark due to different demand dynamics on working and non-working days.

@@

A simple benchmark model is created by fitting a linear regression model to each building and period of the day. The demand for a building at time $t$ is given by
$$
\log y_{t} = \beta_{0,p} + \beta_{1,p} w_{0,t} + \epsilon_{t}, \quad \epsilon_{t} \sim N(0,\sigma^2_{p}),
$$
where $p$ is the 15-minute period of the day at time $t$, $w_{0,t}$ is the scaled temperature experienced^[The 0 subscript denotes no lag and is consistent with Table \@ref(tab:weather-vars) with the building subscript dropped.] at time $t$ and $\epsilon_{t}$ is the residual. We call these our "Individual Linear Regression" (ILR) models.

@@

\begin{table}[t]
\caption{\label{tab:weather-vars} Predictor variables evaluated during feature selection when determining $\mathcal{P}_t$.}

@@

We do not estimate a separate smooth relationship between weather variables and demand for each building in the pooled model. Instead, we estimate the population's relationship. So, for the $i$^th^ predictor we construct a smooth function $f_{i,p}$ for all buildings instead of a set of smooth functions $f_{b,i,p}$ for each building.

@@

As with the PR model, we include subscripts $b$ to denote each building. Unlike the individual formulations, which had separate models fit to each building, all buildings are used when training the mixed effects models and so we include an additional subscript to denote this. Again, to capture changing demand characteristics across the day, separate models are fit for each 15-minute period of the day giving 96 models for each mixed effects formulation.

@@

We use the pooled linear regression model for feature selection as it is quick to fit using ordinary least squares and allows us to model buildings by using a dummy variable for each. Using a linear model also has the advantage of allowing us to efficiently compute the leave-one-out cross-validation (LOOCV) scores [@Seber2012-gu]. Training data is comprised of business days within a window of `r fcst_train_window` days prior to the month we wish to select variables for. As our experimental setup for the validation phase involves one-day ahead forecasts, this variable selection prevents us from using any data from the future. For example, when forecasting for any date in January, only data from the months preceding January would have been used to select predictors.

We use best subset selection during feature selection. Given $p$ predictors we choose the combination of these that produce the best $R^2$ scores. Once the best model based on $R^2$ has been determined for each set of $p$ predictors, we use the LOOCV score to determine the overall best. The LOOCV score is chosen as it gives an estimate of the out of sample performance of our models. Figure \@ref(fig:fs-model-selection) shows the LOOCV scores for each predictor set of size $p$. During feature selection we chose to avoid greedy approaches such as forward or backward stepwise selection; or approaches that work systematically through lagged weather variables [@Hyndman2010-ui]. Naturally, greedy methods have computational benefits, but it is interesting to observe which features are chosen when _all_ possible predictor combinations are assessed.

@@

<!-- It is difficult to observe clear differences between the extended summer and winter periods due to the amount of variance in the chosen variables. However, the maximum temperature over the last 24 hours seems to be more consistently selected during business hours in summer. This seems reasonable as high temperatures in summer will result in additional office cooling. -->

@@

(ref:fs-alpha) Feature selection for summer and winter including shoulder months. Each variable has been coloured based on the type of variable. Scaled temperature is automatically included.

@@

Table \@ref(tab:accuracy-results) shows the MAE, MAPE, sMAPE and MASE for each model across the entire day, during business hours (7:00 am to 7:00 pm) and during non-business hours (7:00 pm to 7:00 am). The INS models have the worst forecasting accuracy across all metrics. This is likely due to overfitting - the natural splines are too flexible for some of the buildings with the amount of data used for training. Evidence of this can be seen in Figure \@ref(fig:one-day-ahead-forecast) where erratic spikes in the predicted demand can be seen. When moving to a mixed effects model that uses natural splines the predictive accuracy is greatly improved. The SSC and SSCATTR models consistently outperform all other models across all metrics.

@@

(ref:one-day-ahead-forecast-cap) One-day ahead forecasts for `r format(p_one_day_ahead_date, "%d %B, %Y")`. The dashed black line shows the actual observed values for this day. The INS models often show erratic jumps due to overfitting.

```{r one-day-ahead-forecast, fig.cap="(ref:one-day-ahead-forecast-cap)", fig.height=8*1.4}
model_df %>%
  filter(fcst_date == p_one_day_ahead_date) %>%
  unnest(preds) %>%
  # filter(bid %in% sample(unique(bid), 3),
  #        fcst_date %in% (dmy("13/3/2017")+0:4)) %>%
  filter(bid %in% sample(unique(bid), 6)) %>%
  mutate(ts = as_datetime(fcst_date) + minutes((period - 1) * 24 * 60 / 96)) %>%
  ggplot(aes(x = ts)) +
  geom_line(aes(y = wh), colour = "black", linetype = "dashed") +
  geom_line(aes(y = pred, colour = model), alpha = 0.7) +
    facet_wrap(~bid, ncol = 2) +
  labs(
    y = expression(Normalised ~ electricity ~ (Wh / m^2)),
    x = "Date",
    colour = "Model"
  ) +
  scale_x_datetime(date_labels = "%H:%M") +
  theme(legend.position = "bottom")
```

@@

Having confirmed that our mixed models produce satisfactory predictions compared to individual models provides us with justification for using them for scenario analysis. We can assess how changing certain variable values increases or decreases energy demand in buildings.

@@

We fit models to `r n_buildings` buildings across Australia. Separate models for each building were fitted as a benchmark. The overall predictive power of several mixed effects models were assessed against this benchmark. One-day ahead forecasts were produced for business days over a year using all forecast methods. Based on the MAE, MAPE, sMAPE and MASE scores of each model the SSC and SSCATTR models performed best. We concluded that moving towards a mixed model approach improved predictive power and also had the additional advantage of allowing us to include fixed effects which could be used for scenario analyses.

@@

```{r fs-all-periods-months, fig.cap="(ref:fs-all-periods-months)", fig.height = 8*1.4}
fs_df %>%
  select(period, month, vars) %>%
  unnest() %>%
  mutate(
    selected = TRUE,
    month = month(month, label = TRUE)
  ) %>%
  spread(vars, selected, fill = FALSE) %>%
  # Ensure models with no variables selected have white squares plotted
  right_join(list(
    month = unique(month(fcst_test_dates)),
    period = fcst_test_periods
  ) %>%
    cross_df() %>%
    mutate(month = month(month, label = TRUE)),
  by = c("month", "period")
  ) %>%
  gather(vars, selected, -c(period, month)) %>%
  mutate(
    selected = tidyr::replace_na(selected, FALSE),
    vars = fct_rev(factor(vars, levels = predictor_vars)),
    predictor_type = case_when(
      str_detect(vars, "^scaled_temperature$") ~ "Forced variable",
      str_detect(vars, "^scaled_temperature_lag") ~ "Temperature lags",
      str_detect(vars, "^scaled_temperature_(?!lag)") ~ "Temperature statistic",
      str_detect(vars, "^scaled_wh_lag") ~ "Demand lags",
      TRUE ~ "Other"
    )
  ) %>%
  filter(selected == TRUE) %>%
  ggplot(aes(x = period, y = vars, fill = predictor_type)) +
  geom_tile() +
  facet_wrap(~month) +
  # scale_fill_manual(values = c("white", "#3d6b78")) +
  labs(
    x = "Period",
    y = "Predictor",
    fill = "Predictor type"
  ) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ) +
  scale_x_continuous(breaks = seq(8, 96, 8))
```
\newpage
